---
title: "Exercises G"
params:
  answers: true
author: "Gerko Vink"
date: "Data Science and Predictive Machine Learning"
output: html_document
---

# Introduction

---

In this practical, we will focus on two different classification methods: K-nearest neighbours and logistic regression.

One of the packages we are going to use is `class`. For this, you will probably need to `install.packages("class")` before running the `library()` functions. `ISLR` is also a new package, that needs to be installed to access the `Default` data.

```{r packages, warning = FALSE, message = FALSE}
library(MASS)
library(magrittr)
library(class)
library(ISLR)
library(tidyverse)
library(caret)
```

Before starting with the exercises, it is a good idea to set your seed, so that (1) your answers are reproducible and (2) you can compare your answers with the answers provided. 

```{r seed}
set.seed(123)
```

---

# Default dataset from package `ISLR`

---

The default dataset contains credit card loan data for 10 000 people. The goal is to classify credit card cases as `yes` or `no` based on whether they will default on their loan.

---

1. __Create a scatterplot of the `Default` dataset, where `balance` is mapped to the x position, `income` is mapped to the y position, and `default` is mapped to the colour. Can you see any interesting patterns already?__ 

---

```{r defaultplot1, include = params$answers}

```

https://www.r-bloggers.com/2021/05/lasso-regression-model-with-r-code/


---

End of Practical


