---
title: "Practical I"
author: "Gerko Vink"
date: "Data Science and Predictive Machine Learning"
output: 
   html_document:
    toc: true
    toc_depth: 5
    toc_float: true
    number_sections: false
---
  
<style type="text/css">
  
body{ /* Normal  */
  font-size: 12px;
  }
td {  /* Table  */
  font-size: 12px;
}
h1.title {
  font-size: 18px;
  color: DarkBlue;
}
h1 { /* Header 1 */
  font-size: 18px;
}
h2 { /* Header 2 */
  font-size: 18px;
}
h3 { /* Header 3 */
  font-size: 18px;
}
code.r{ /* Code block */
  font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
  font-size: 14px;
}
</style>

We use the following packages:
```{r packages, warning = FALSE, message = FALSE}
library(tidyverse)
library(magrittr)
library(caret)
library(kernlab)
library(MLeval)
library(pROC)
```

# Introduction
Let's take the `titanic` data that we used before and fit the following four models on a training version (70% of cases) of that data set.

1. A logistic regression model
2. A linear kernel SVM
3. A polynomial kernel SVM
4. A radial kernel SVM

Finally, compare the performance of all 4 techniques on the test version (30% of not yet used cases) of that data set. 

---

# Grab the data
We can use the following code block to directly load the data in our workspace:
```{r}
con <- url("https://www.gerkovink.com/erasmus/Day%202/Part%20D/titanic.csv")
titanic <- read_csv(con)
```

---

# Prepare the data
We need to take care of some columns that are not well-coded. Let's make all the measurement levels as they are supposed to be. That means factors into factors, ordered factors into ordered factors, etc. 
```{r}
titanic %<>% 
  mutate(Pclass   = factor(Pclass, 
                         ordered = TRUE, 
                         labels = c("1st class", "2nd class", "3rd class")), 
         Survived = factor(Survived, 
                           labels = c("Died", "Survived")))

str(titanic)
```
The `%<>%` pipe returns the result of the pipe to the object. 

--- 

# Exercise
Fit the four above mentioned models on the training set and evaluate their performance on the test set. 

---


End of Practical